---
layout: page
author: ricardo mayer
---

# Introducing a basic ('toy') model #


In theory, changes in productivity can have counter-intuitive effects. For instance, as noted by xxx, yyy and more recently zzz and aaa, if productivity has a long term growth rate of 2\% and suddenly rises to a new long term growth rate of 4\%, which is basically very good news, a recession ensues. The basic intuition is that those news are even better than a permanent gain in productivity, it's a permanent increase in the speed of productivity growth and people feel they can afford to work and invest less than before, because even if we don't have that much capital or don't put in much hours, salaries and capital rental payments are going to be higher sooner than before and they can afford to eat more and work less.

To fix ideas, let's see what it means for productivity to grow at 2, 3 or 4 percent every year. If it grows at 2\%, it will take 35 years to double its size. If it grows at 3\% or 4\% it will only take only 23 or 17 years to double, respectively. In the same period that it takes to double its size by growing at 2\%, productivity would have almost tripled if growing at 3\% or it would be 4 times bigger if growing at 4\%. Those are significant changes induced by a rise of the long term growth. If wages and rental price of capital rise with productivity, these are potentially large changes in wealth, too.

Some researchers like rebelo and jaimovich focus their attention on the pure-income effect on labor supply and devise a type of utility function that make agents to less likely to work less just because future income now seems higher.The problem  to be addressed was oversensitive preferences for leisure. But I'm interested in different line of attack: changes in trend growth may occur, but may fail to act on them due to informational imperfections. Maybe the cannot tell for sure if they are facing a temporary or a permanent change in trend growth. Maybe even if they were sure the best approximation to what is happening is a permanent long term growth, the are unsure about possible deviations of this behavior that they cannot anticipate or formulate. The former deals with imperfect information and the second with methods of robust control. In any case, the permanent good news are treated with more caution (it can be temporary, it can evolve in unexpected ways) and as a result we don't see people working and investing less. In this line, the problem to be addressed is full information or excess of certainties.

In what follow, we will build, discuss and solve the simplest environment that can think of preferences, productivity and technology that will allow me to examine the consequences of permanent changes in trend growth under full information, imperfect information and unstructured model uncertainty. I will grant the not so gracious name of Toy Model to such environment and its setup is described in the next section.

Stop here if you don't need or don't want to review the differences in the evolution of productivty as the result of permanent, persistent and transitory  changes in growth rates. Otherwise, check the numerical examples below. 


## Further numerical examples on trend growth ##

Let's keep the focus on growth rates of 2\% and 4\% and doubling the current level of productivity. I'll use everywhere the logarithmic approximation to growth rates: \( \dfrac{x_t - x_{t-1}}{x_{t-1}} \approx \ln x_t - \ln x_{t-1}  \) which are good approximations for the small growth rates I'm  considering 

### Question: How different are trajectories at constant growth rates of 2 and 4 per cent? ###
The direct approach would be to plot the level of productivity for each case, which corresponds to plot the function \( A_t = e^{\mu t} A_0 \) for \( \mu = 0.02  \) and \( \mu = 0.04\). However, I like to plot also the logartihm of productivity: \( \ln A_t = \mu t + \ln A_0 \) plots are no obscured by too high values, curves with the same growth rate are parallel and easy to spot and the math is slightly simpler. 

Here is the plot in logs, where I also added a line at \(\log(2)\) which represent the double of the initial level of productivity (\(A_0=1, ~ \ln A_0=0\)).

```{r growth_2_4_log_plot, fig.align='center', echo=FALSE}
finalT = 40
xvals = 0:finalT
level.start = 1
level.target = 2*level.start
base.rate = 0.02
new.rate = 0.04

rates_base = c(0,rep(base.rate, finalT))
rates_new =  c(0,rep(new.rate, finalT))

values.log.baserate = cumsum(rates_base) + log(level.start)
values.log.newrate = cumsum(rates_new) + log(level.start)

ts_log.baserate = as.ts(values.log.baserate, start=0)
ts_log.newrate = as.ts(values.log.newrate, start=0)
plot(cbind(ts_log.baserate,ts_log.newrate),
     plot.type="single")
abline(h=log(level.target) )
```



Now, the plot for levels (i.e. productivity level as an exponential function)
```{r growth_2_4_level_plot, fig.align='center', echo=FALSE}
plot(exp(cbind(ts_log.baserate,ts_log.newrate)),
     plot.type="single")
abline(h=level.target )
```


While is clear that the 4\% curve hits the doubling mark sooner than the 2\% curve, it's not so clear at which period the productivity has doubled. That's our next question.

### Question: How long it takes to double productivity for growth rates of 2 and 4 percent? ###

For constant growth rates it's easy to give an exact answer: 

\[ t = \frac{\ln(2)}{\mu}, \Rightarrow t = 34.66 \text{ if } \mu=0.02, \quad  t = 17.33 \text{ if } \mu = 0.04\]

Thus, 34 periods are not enough to double in the first case and need a little more than 17 periods in the second. Thus the first integer date when you have doubled productivity will be 35 or 18, depending on the growth rate. If 3\% is our base rate for comparision, 35 periods will be our time-lapse of reference.

The preceeding formula allows to plot this time-to-doubling against growth rates:
```{r time-to-doubling, fig.align='center', echo=FALSE}
gr_rates = seq(0.01,0.06,0.001)
time_to_double = log(2)/gr_rates
plot(gr_rates,time_to_double, type="l")
```


A less elegant solution is to just write a code that will return the date of first ocurrence of a productivity greater than double of the starting level. It is not as neat as the formula above but it's handy for asking this same question when growth rates are arbitrary sequences.

Here is an example for the base growth rate of 2\%:

```{r doubling-time-base-rate}
double_period_base_rate = which.max(values.log.baserate >= (log(2)+log(level.start))) -1
double_period_base_rate
```

### Question: how different is a transitory change to 4\% than a permanent one ###
To be precise, we mean by transitory that it only affects the growth rate in period 1, thus it will grow at 4\% in period one and 2\% from thereafter. Doubling comes slightly faster than in the baseline case of 2\% but very far from the permanent 4\%: it takes 34 periods to double, compared with the 35 and 18 we saw before.

I particulary like how simple is the plot in logs: it copies the trayectory of the 4\% curve for a short while (one period) and then follows a parallel line to the 2\%, rather close to the baseline curve:

```{r transitory_vs_perms_log_plot, fig.align='center', echo=FALSE}
diff.rate =  new.rate -base.rate 
rates_tempnew = rates_base
rates_tempnew[2] = new.rate
values.log.tempnewrate = cumsum(rates_tempnew) + log(level.start)
double_period_tempnew_rate = which.max(values.log.tempnewrate >= (log(2)+log(level.start))) -1
ts_log.tempnewrate = as.ts(values.log.tempnewrate, start=0)
plot(cbind(ts_log.baserate,ts_log.newrate, ts_log.tempnewrate),
     plot.type="single")
abline(h=log(level.target) )
```



### How different are (very) persistent but eroding changes from transitory and permanent ones? ###
For very persisting changes I mean that this increase (from 2\% to 4\% in period one) has an autorregressive coefficient of 0.8, 0.9 or 0.99, and you can see the results in the following plot that adds log-trayectories of this three different persistances to our last plot (yes, six rays are a lot, I may take out the 0.90 case):

```{r persistent_vs_perms_log_plot, fig.align='center', echo=FALSE}
persistence80 = 0.8
persistence90 = 0.9
persistence99 = 0.99

rates_80_coeffs = c(c(0,1), cumprod(rep(persistence80,finalT-1)) )
rates_90_coeffs = c(c(0,1), cumprod(rep(persistence90,finalT-1)) )
rates_99_coeffs = c(c(0,1), cumprod(rep(persistence99,finalT-1)) )

rates_80 = rates_base + (diff.rate*rates_80_coeffs)
rates_90 = rates_base + (diff.rate*rates_90_coeffs)
rates_99 = rates_base + (diff.rate*rates_99_coeffs)

values.log.persistence_80 = cumsum(rates_80) + log(level.start)
values.log.persistence_90 = cumsum(rates_90) + log(level.start)
values.log.persistence_99 = cumsum(rates_99) + log(level.start)

double_period_persistence_80 = which.max(values.log.persistence_80 >= (log(2)+log(level.start))) -1
double_period_persistence_90 = which.max(values.log.persistence_90 >= (log(2)+log(level.start))) -1
double_period_persistence_99 = which.max(values.log.persistence_99 >= (log(2)+log(level.start))) -1

ts_log.persistence80 = as.ts(values.log.persistence_80, start=0)
ts_log.persistence90 = as.ts(values.log.persistence_90, start=0)
ts_log.persistence99 = as.ts(values.log.persistence_99, start=0)

plot(cbind(ts_log.baserate,ts_log.newrate, ts_log.tempnewrate,
           ts_log.persistence80, ts_log.persistence90,  ts_log.persistence99),
     plot.type="single")
abline(h=log(level.target) )
```
For autorregressive coefficients of 0.8, 0.9 and 0.99 doubling times are 30, 26 and 19 periods, respectively. As you can see there is a lot of differences between 0.8 and 0.99, but 0.8 is still very significant compared to the pure transitory change.



### How large has to be a transitory change to growth rate to reach the doubling mark as fast as a persitent change? ###

The temporary change in growth would need to take the baseline rate of 2\% to 12\%, 20\% and 34\% to reach the doubling point as fast as the 0.8, 0.9 and 0.99 persistence cases, respectively. And it will have to increase temporarily the growth rate up to 36 \% to mimic the permanent change in this aspect.

I'll show you different matching comparison when we are ready to compute discounted sums of log productivities.


### After 35 periods how large is producitivity under all of these different scenarios? ###

After 35 periods growing at 2\% productivity increased by a factor of 2.01, which is not so different of what we have with our temprary change: the factor is 2.05. The three persistent cases have factors of 2.22, 2.44 and 3.64.

This would be so much better with a table. TODO.
